### 2024-05-21 thesis/input_files/literaturereview.tex
May 13th 2024
corrected literature review

%Electricity Forecasting 
%intro
During the past 25 years %because paper of 2014 says 15, so now it is 25
a wide range of new ideas have been proposed for electricity point forecasting and for probabilistic forecasting.
\\
The field benefitted greatly from the increse of computing power, the greater availability of data and the interest in data science.
As a consequence, the forecaster's toolbox has grown in size and complexity.
\\
Before delving into the literature review, we stress that at this point in time there is no superior method. Different solutions may outperform or underperform compared to other techniques depending on the problem settings. Thus, understanding the complexity, strenghts and weaknesses of each method is crucial for fitting the right model to the right setting.
\\
Within this research community, it emerged the need for more homogeneity in the choice of the error valuation metrics (Section \ref{metrics}), data quality and in the way of comparing model performances \cite{EPF_review}. As a solution, \cite{lago} proposes a checklist to aid evaluating the meaningfulness of new research. Throughout this thesis work, we will stick to the proposed principles and best practices peculiar of the EF field.
%literature review

@@

Electricity forecasting is a vague term and it is used in the literature to refer to the whole field. Thus, in order to introduce some clarity it is useful to classify the range of EF articles in terms of their core attributes.

@@

In terms of forecasting horizons, we can group EF into four major categories: very short term forcasting (VSTF), short term load forecasting (STF), medium term forecasting (MTF) and long term forecasting (LTF). Consensus in the literature is to use one day, two weeks and three years respectilvely \cite{hong_phd} as cut off horizons; see \ref{fig:time} for a visualisation. 

@@

  \caption{Time classsification \cite{prob_elf}}

@@

EF literature distinguishes between point and probabilistic forecasts. Each of the two has its advantages and disadvantages. Point forecasts are easier to generate and less computationally intensive while probabilistic forecast are more informative. Industry and research efforts have focused primarily on point forcasting. Nevertheless, interest in probabilistic forecasting has risen considerably over the last years due to renewable integration requirements, introduction of smart grids and increased market competitivness.

@@

  \caption{Point versus probabilistic pubblications over the past years}

@@

  \caption{Pubblications by method over the past years}

@@

EF is a heterogenous field of research, its researchers come from a wide array of backgrounds, with electrical engineers and statisticians making up the top contributors; their different educational training may explain why the split between statistical and computational intelligence methods is so marked. Figure \ref{fig:subject_area} depitcs the EF publications by subject area. What can be concluded, is that the bulk of pubblication come from engineering, computer science, mathematics and econometrics.

@@

  \caption{Pubblications by subject area}

@@

\section{Kernel methods literature review}
%INTRO KERNEL METHODS E ENERGY PRICE MODELLING
Kernel methods are a class of algorithms for pattern analysis.
With kernel methods we are able to apply linear methods with predictors in a high dimensional space, without having to explicitly evaluate the involved dot products of the features.
Throughout this thesis work, we will address the performance of kernel methods in the context of EF.
\\
%- Kernel history evolution.
Their name comes from the German word kern, which translates to core in english. Such term was first used by David Hilbert in his paper on integral equations \cite{hilbert} where he introduced the term "definite kernels". Following, Hilbert's and Schmidt's work \cite{schmidt} lead to the introduction of a new space, the Hilbert space.
In 1909, James Mercer improved Hilbert's work by proposing his theorem \cite{mercer}. This theorem underlies the power of kernel methods, that is the kernel trick.
In 1938, Schoenberg \cite{schoenberg} developed the mathematical results that allow us to find the kernel associated to a specific feature space metric.
In 1941, Kolmogorov \cite{kolmogorov} carried out stuides on representing kernel in linear spaces.
In 1950, Aronszajn \cite{aronszajn} published the first work on reproducing kernel hilbert spaces (RKHS); developing the general method for representing kernels in linear spaces.
In 1964, Aizerman \cite{aizerman} further improved the theory of RKHS.
It was in the nineties that theory of kernel methods got popular, particularly in the field of machine learning. Kernels have been used in various different tasks such as SVM \cite{vapnik1} \cite{vapnik2}, Gaussian process classifiers \cite{williams}, spline methods \cite{wahba}, neural networks \cite{poggio} and principal component analysis \cite{pca_scholkopf}.
Nevertheless kernel methods received very little attention in the specific setting of EF literature.
\\
The kernel theory needed for this thesis work is covered in Section \ref{kernel_theory}. 
For an introduction to kernel methods, we refer to \cite{learning_with_kernels, hofmann2006review}.
\\
Kanagawa et Fukumizu introduces to the concept of kernel mean embedding \cite{pmlr}. Muandet et al. \cite{Muandet_2017} surveys established results and new advances in the theory of Hilbert space distribution embeddings. 
It has to be said that, computing and storing such embeddings becomes prohibitive for large scale settings. Rudi et al. \cite{2022nystrom} proposes an efficient approximation procedure based on the Nyström method \cite{nystrom}, providing also an upper bound for the approximation error.
Smola et al. \cite{supersamples} presents kernel herding; basically, the authores used the kernel trick to extend the herding algorithm to continous spaces. The result is an infinite memory deterministic process that takes in a collection of samples and learns to approximate a probability density function (PDF).
\\
%QUESTO DA VEDERE
% Particularly extending the idea of \cite{2022nystrom}, where the Nyström approximation is employed in computing the kernel mean embedding, experiments with the Pivoted Cholesky decomposition will be performed.

>>>>

%Electricity Forecasting 
%intro
During the past 25 years %because paper of 2014 says 15, so now it is 25
a wide range of new ideas have been proposed for electricity point forecasting and for probabilistic forecasting.
\\
The field benefitted greatly from the increse of computing power, the greater availability of data and the interest in data science.
As a consequence, the forecaster's toolbox has grown in size and complexity.
\\
Before delving into the literature review, we stress that at this point in time there is no superior method. Different solutions may outperform or underperform compared to other techniques depending on the problem settings. Thus, understanding the complexity, strenghts and weaknesses of each method is crucial for fitting the right model to the right setting.
\\
Within this research community, the need for more homogeneity in the choice of the error valuation metrics emerged (Section \ref{metrics}), data quality and in the way of comparing model performances \cite{EPF_review}. As a solution, \cite{lago} proposes a checklist to aid evaluating the meaningfulness of new research. Throughout this thesis work, we will stick to the proposed principles and best practices peculiar of the EF field.
%literature review

@@

Electricity forecasting is a vague term and is used in the literature to refer to the whole field. Thus, in order to introduce some clarity, it is useful to classify the range of EF articles in terms of their core attributes.

@@

In terms of forecasting horizons, we can group EF into four major categories: very short term forcasting (VSTF), short term load forecasting (STF), medium term forecasting (MTF) and long term forecasting (LTF). Consensus in the literature is to use one day, two weeks and three years respectilvely \cite{hong_phd} as cut off horizons; see figure \ref{fig:time} for a visualisation. 

@@

  \caption{Time classification \cite{prob_elf}}

@@

EF literature distinguishes between point and probabilistic forecasts. Each of the two has its advantages and disadvantages. Point forecasts are easier to generate and less computationally intensive while probabilistic forecasts are more informative. Industry and research efforts have focused primarily on point forcasting. Nevertheless, interest in probabilistic forecasting has risen considerably over the last years due to renewable integration requirements, introduction of smart grids and increased market competitiveness.

@@

  \caption{Point versus probabilistic publications over the past years}

@@

  \caption{Publications by method over the past years}

@@

EF is a heterogenous field of research, its researchers come from a wide array of backgrounds, with electrical engineers and statisticians making up the top contributors; their different educational training may explain why the split between statistical and computational intelligence methods is so marked. Figure \ref{fig:subject_area} depitcs the EF publications by subject area. What can be concluded, is that the bulk of publication come from engineering, computer science, mathematics and econometrics.

@@

  \caption{Publications by subject area}

@@

\section{Kernel methods literature review}
%INTRO KERNEL METHODS E ENERGY PRICE MODELLING
Kernel methods are a class of algorithms for pattern analysis.
With kernel methods we are able to apply linear methods with predictors in a high dimensional space, without having to explicitly evaluate the involved dot products of the features.
Throughout this thesis work, we will address the performance of kernel methods in the context of EF.
\\
%- Kernel history evolution.
Their name comes from the German word kern, which translates to core in english. Such term was first used by David Hilbert in his paper on integral equations \cite{hilbert} where he introduced the term "definite kernels". Following, Hilbert's and Schmidt's work \cite{schmidt} lead to the introduction of a new space, the Hilbert space.
In 1909, James Mercer improved Hilbert's work by proposing his theorem \cite{mercer}. This theorem underlies the power of kernel methods, that is the kernel trick.
In 1938, Schoenberg \cite{schoenberg} developed the mathematical results that allow us to find the kernel associated to a specific feature space metric.
In 1941, Kolmogorov \cite{kolmogorov} carried out stuides on representing kernel in linear spaces.
In 1950, Aronszajn \cite{aronszajn} published the first work on reproducing kernel hilbert spaces (RKHS); developing the general method for representing kernels in linear spaces.
In 1964, Aizerman \cite{aizerman} further improved the theory of RKHS.
It was in the nineties that theory of kernel methods got popular, particularly in the field of machine learning. Kernels have been used in various different tasks such as SVM \cite{vapnik1} \cite{vapnik2}, Gaussian process classifiers \cite{williams}, spline methods \cite{wahba}, neural networks \cite{poggio} and principal component analysis \cite{pca_scholkopf}.
Nevertheless kernel methods received very little attention in the specific setting of EF literature.
\\
The kernel theory needed for this thesis work is covered in Section \ref{kernel_theory}. 
For an introduction to kernel methods, we refer to \cite{learning_with_kernels, hofmann2006review} and \cite{shawe2004kernel}.
\\
Kanagawa et Fukumizu introduces to the concept of kernel mean embedding \cite{pmlr}. Muandet et al. \cite{Muandet_2017} surveys established results and new advances in the theory of Hilbert space distribution embeddings. 
It has to be said that, computing and storing such embeddings becomes prohibitive for large scale settings. Rudi et al. \cite{2022nystrom} proposes an efficient approximation procedure based on the Nyström method \cite{nystrom}, providing also an upper bound for the approximation error.
Smola et al. \cite{supersamples} presents kernel herding; basically, the authores used the kernel trick to extend the herding algorithm to continous spaces. The result is an infinite memory deterministic process that takes in a collection of samples and learns to approximate a probability density function (PDF).
\\
%QUESTO DA VEDERE
% Particularly extending the idea of \cite{2022nystrom}, where the Nyström approximation is employed in computing the kernel mean embedding, experiments with the Pivoted Cholesky decomposition will be performed.
